Subsidiary Recursion in Coq

### 2022-02-03 paper/paper.tex
Additional edits through Section 2.

We begin by summarizing the interface our derivation provides for
subsidiary recursion (Section~\ref{sec:interface}), and we then
present some examples (Section~\ref{sec:examples}).  We next explain
how the interface is actually implemented (Section~\ref{sec:deriv}),
including our definition of retractive-positive recursive types
(Section~\ref{sec:mu}).  We then cover the interface for subsidiary
induction (Section~\ref{sec:interfacei}), and give some example proofs
using it (Section~\ref{sec:examplesi}). We conclude with a discussion
of related work (Section~\ref{sec:related}).

@@

Our approach is within a long line of work using ideas from universal
algebra and category theory to describe inductive datatypes and their
recursion principles (cf.~\cite{traytel12,cockett92,hagino87}).  On this approach, one describes transformations
to be performed on data as \emph{algebras}, which can then be
\textit{folded} over data.  The simplest form of algebras, namely
$F$-algebras for functor $F$, are morphisms from $F\ A$ to $A$, for carrier object
$A$.  From a programming perspective, an $F$-algebra is given input of
type $F\ A$, and must compute a result of type $A$.  An example of $F$
is the signature functor for lists, which we will use below:

\begin{minted}{coq}
Inductive ListF(X : Set) : Set :=
| Nil : ListF X
| Cons : A -> X -> ListF X.
\end{minted}

Algebras for our subsidiary recursion are more complex than
$F$-algebras.  Let us begin with an informal explanation.  For reasons
we will explain further below, the carrier of the algebra will be a
functor \verb|X : Set -> Set|.  The algebra is presented with:

@@

We will use Coq inductive types for the signature functors \verb|F| of
various datatypes, thus enabling recursions to use Coq's
pattern-matching on the subdata structure \verb|d|.  So the style of
coding against this interface retains a similar feel to structural
recursion.  Unlike with structural termination, though, the interface
here is type-based and hence compositional.

We have previously dubbed this interface a \emph{recursion
universe}~\cite{stump20}.  As in other domains using the term
``universe'', we have a kind of space (here, \verb|R|), which one
cannot escape using certain operations.  Other examples are the
ordinal $\epsilon_0$ and $\omega^-$, and the physical universe and
traveling at the speed of light.  Staying in the recursion universe is
good, because we may recurse (via \verb|rec|) on any value of type
\verb|R|.  Some points must still be explained: why \verb|X| has type
\verb|Set -> Set|, and the definition of \verb|FoldT|.  Let us see
these details next.

@@

\begin{figure}
\begin{minted}{coq}
Definition List := Subrec ListF.
Definition inList : ListF List -> List := inn ListF.
Definition mkNil : List := inList Nil.
Definition mkCons (hd : A) (tl : List) : List := inList (Cons hd tl).
Definition toList : list A -> List.
Definition fromList : List -> list A.
\end{minted}
  \caption{Some basics from \texttt{List.v}, specializing the functor-generic derivation of subsidiary recursion to lists (\texttt{List.v})}
  \label{fig:listf}
\end{figure}

The type over which one can recurse with our scheme of subsidiary
recursion is called \verb|Subrec|.  It is parametrized by a signature
functor \verb|F| of type \verb|Set -> Set|.  \verb|Subrec| comes with
\verb|inn : F Subrec -> Subrec|, which behaves computationally like a
constructor.  We will later derive an induction principle for this type
(Section~\ref{sec:interfacei}). The definition of \verb|Subrec| uses
retractive-positive recursive types, to take a fixed-point of a
construction based on \verb|F|.  We present these recursive types in
Section~\ref{sec:mu} below.

For our examples, we will consider the specialization to the case of
lists, parametrized by the type \verb|A| of elements.  In general, to
use our development to get subsidiary recursion over some datatype,
one must define a signature functor for the datatype.  For lists, this
is \verb|ListF|, which we saw at the start of Section~\ref{sec:interface}.  \verb|List| is defined to
be \verb|Subrec|, with \verb|F| instantiated to \verb|ListF A|.  This
type \verb|List| is not to be confused with the type \verb|list| of
lists in Coq's standard library.  As noted previously, our development
is meant to be used in extension of existing inductive datatypes, not
replacing them.  The figure also shows constructors \verb|mkNil| and
\verb|mkCons| for \verb|List|, and typings for conversion functions
between \verb|List| and \verb|list| (code elided).

@@

We need a fixed-point here due to the input \verb|fold| (\verb|Algf|)
of type \verb|FoldT Alg R|.  This is the type for fold functions which
apply algebras (\verb|Alg|) to data of type \verb|R|.  The variable \verb|Alg| occurs only positively (but not
strictly positively) in \verb|AlgF|, because it occurs negatively in
\verb|FoldT Alg R| which occurs negatively in \verb|AlgF Alg X|.  So
we can indeed take a fixed-point of \verb|AlgF| to define the constant
\verb|Alg|.

@@

Let us look at \verb|AlgF|.  As noted already, each recursion is based
on an abstract type \verb|R|, representing the data upon which we will
recurse.  This is the first argument to a value of type
\verb|AlgF Alg X|.  Reasoning parametrically, an algebra can assume
nothing about \verb|R| except that it supports the following
operations.  Next we have a local
\texttt{fold} function, which will allow us to fold another algebra over data
of type \verb|R|.  We will use \verb|fold| to initiate subsidiary
recursions.  Then there is \verb|rec|, for recursive calls on data of
type \verb|R|.

As noted already, for subsidiary recursion, algebras have a carrier
\verb|X| which depends (functorially) on a type.  When we fold an
algebra using a fold function (either global or local) of type
\verb|FoldT Alg C|, (i) recursive calls may compute a result of type
\verb|X R|, mentioning the abstract type \verb|R| for that recursion;
and (ii) outside that recursion, the result will have type \verb|X C|.
Having a functor for the carrier of the algebra gives us the
flexibility to type results inside a recursion with the abstract type
\verb|R|, but view those results as having the type \verb|C| outside
the recursion.  The function \verb|fold| in the figure initiates
top-level folds.  We also can have functions
between \verb|Alg| and its \verb|Algf|-unfolding.  We will
return to the code for \verb|Subrec.v| in Section~\ref{sec:deriv}.

>>>>

We begin by summarizing the interface our derivation provides for
subsidiary recursion (Section~\ref{sec:interface}), and we then
present some examples (Section~\ref{sec:examples}).  We next explain
the implementation of our interface (Section~\ref{sec:deriv}),
including our definition of retractive-positive recursive types
% (Section~\ref{sec:mu})
.  We then cover the interface for subsidiary
induction (Section~\ref{sec:interfacei}), and give some example proofs
that use it (Section~\ref{sec:examplesi}). We conclude with a discussion
of related work (Section~\ref{sec:related}).

@@

Our approach is within a long line of work using ideas from universal
algebra and category theory to describe inductive datatypes and their
recursion principles (cf.~\cite{traytel12,cockett92,hagino87}). Under
this approach, inductive datatypes are equipped with a signature
functor which explains how to 'unroll' one layer of the type. The
signature functor for lists is:
\begin{minted}{coq}
Inductive ListF (X : Set) : Set :=
| Nil : ListF X
| Cons : A -> X -> ListF X.
\end{minted}
Transformations on data with functor $F$ are described as
\emph{algebras} which are then \textit{folded} over data.  The
simplest form of algebras are morphisms from $F\ A$ to $A$, for some
carrier object $A$.  From a programming perspective, an $F$-algebra is
given input of type $F\ A$, and must compute a result of type $A$. The
$ListF$-algebra for computing the length of a list is:
\begin{minted}{coq}
Definition LengthAlg (l : ListF nat) : nat :=
match l with
| Nil => 0
| Cons a len' => 1 + len'
end.
\end{minted}
Algebras for our subsidiary recursion are more complex than
$F$-algebras.  Let us begin with an informal explanation.  For reasons
we will explain further below, the carrier of the algebra will be a
functor \verb|X : Set -> Set|.  Our algebras are presented with:

@@

We use Coq inductive types for the signature functors \verb|F| of
various datatypes. This allows algebras to use Coq's
pattern-matching on the subdata structure \verb|d|, enabling a style
of coding that retains a similar feel to structural recursion. Unlike
structurally recursive functions, however, this interface uses a
type-based (and hence compositional) criteria to ensure termination.

We have previously dubbed this interface a \emph{recursion
  universe}~\cite{stump20}.  As in other domains using the term
``universe'', we have a kind of space (here, \verb|R|), which one
cannot escape using the provided operations.  Other examples include the
ordinal $\epsilon_0$ and operator $\omega^-$, and the physical
universe and traveling at the speed of light.  Staying in the
recursion universe is good, because we may recurse (via \verb|rec|) on
any value of type \verb|R|.  Some points must still be explained: in
particular why \verb|X| has type \verb|Set -> Set|, and the definition
of \verb|FoldT|. We consider those details next.

@@

\begin{figure}
\begin{minted}{coq}
Definition List := Subrec ListF.
Definition inList : ListF List -> List := inn ListF.
Definition mkNil : List := inList Nil.
Definition mkCons (hd : A) (tl : List) : List := inList (Cons hd tl).
Definition toList : list A -> List.
Definition fromList : List -> list A.
\end{minted}
  \caption{Some basics from \texttt{List.v}, specializing the
    functor-generic derivation of subsidiary recursion to
    lists parameterized by an element type \texttt{A}  (\texttt{List.v})}
  \label{fig:listf}
\end{figure}

The type over which one can recurse using our subsidiary
recursion scheme is called \verb|Subrec|.  It is parametrized by a signature
functor \verb|F| of type \verb|Set -> Set|.  \verb|Subrec| comes with
\verb|inn : F Subrec -> Subrec|, which behaves computationally like a
constructor.  We will later derive an induction principle for this type
(Section~\ref{sec:interfacei}). The definition of \verb|Subrec| uses
retractive-positive recursive types, to take a fixed-point of a
construction based on \verb|F|.  We present these recursive types in
Section~\ref{sec:mu} below.

Our examples specialize this scheme to the case of polymorphic lists,
presented in Figure~\ref{fig:listf}.  In general, to use our
development to get subsidiary recursion over some datatype, one must
define a signature functor for the datatype, e.g., \verb|ListF| from
the start of this section. We define \verb|List| to be
\verb|Subrec| with \verb|F| instantiated to \verb|ListF A|.  This
type \verb|List| is not to be confused with the type \verb|list| of
lists from Coq's standard library.  As noted previously, our development
is meant to be used in extension of existing inductive datatypes, not
replacing them.  The figure also shows constructors \verb|mkNil| and
\verb|mkCons| for \verb|List|, and typings for conversion functions
between \verb|List| and \verb|list| (we elide their definitions).

@@

We need a fixed-point here due to the \verb|fold| parameter
\verb|AlgF| expecting an argument of type \verb|FoldT Alg R|.  This is the type for fold functions which
apply algebras (\verb|Alg|) to data of type \verb|R|.  The variable \verb|Alg| occurs only positively (but not
strictly positively) in \verb|AlgF|, because it occurs negatively in
\verb|FoldT Alg R| which occurs negatively in \verb|AlgF Alg X|.  So
we can indeed take a fixed-point of \verb|AlgF| to define the constant
\verb|Alg|.

@@

Let us look at \verb|AlgF|.  As noted already, each recursion is based
on an abstract type \verb|R|, representing the data upon which we will
recurse.  This is the first argument to a value of type
\verb|AlgF Alg X|.  Reasoning parametrically, an algebra can assume
nothing about \verb|R| except that it supports the
\verb|fold| and \verb|rec| operations that follow. The first is a local
\verb|fold| function, which allows us to fold another algebra over data
of type \verb|R|.  We will use \verb|fold| to initiate subsidiary
recursions.  The second, \verb|rec|, is used to make recursive calls
on values of type \verb|R|.

As noted already, for subsidiary recursion, algebras have a carrier
\verb|X| which depends (functorially) on a type.  When we fold an
algebra using a fold function (either global or local) of type
\verb|FoldT Alg C|, (i) recursive calls may compute a result of type
\verb|X R|, mentioning the abstract type \verb|R| inside the recursion;
and (ii) outside that recursion, the result will have type \verb|X C|.
Having a functor for the carrier of the algebra gives us the
flexibility to type results inside a recursion with the abstract type
\verb|R|, but view those results as having the type \verb|C| outside
the recursion.  The function \verb|fold| in Figure~\ref{fig:algf} initiates
top-level folds.  We can also have functions
between \verb|Alg| and its \verb|AlgF|-unfolding.  We will
return to the code for \verb|Subrec.v| in Section~\ref{sec:deriv}.
